Basic concepts and tools for the Toki Pona minimalist language: Wordnet synsets; analysis, synthesis and syntax highlighting of texts

### 2017-12-01 article/article.tex
andando com stats

\subsection{Statistics of the vocabulary}
In~\cite{pije} are statistics about Toki Pona corpus.
This section focuses on the statistics of the vocabulary
and syntactic rules:
the letters, phonemes, word sizes,
possible combinations for words and sentences.
The statistics are in Appendix~\cite{listings},
and the next paragraph is an overview.
Python scripts were used to obtain the measurements
and are available at~\cite{tokipona}.

@@

Vowel frequencies are as shown in Table~\ref{freqLet}
for starting, end and internal position.
Consonants are shown in Table~\ref{tab:cons}
for meaningful values.
Comparison of vowels agains consonant is in Table~\ref{tab:vcon}.
The high incidences of 'l' and 'a' favor
the naturalness of the language prosody
as it resembles lallation.

@@

Most often phonemes are in
Table~\ref{tab:phon}.
A toolbox for obtaining such statistics and sets
of words is in~\cite{tokipona}.

Given the rules given in Section~\ref{phonology}
(of 15 letters, 5 vowels, (C)V(N) phonemes,
forbidden ji, wu, wo, ti, nn, nm),
X words are possible with 1 syllable,
Y with 2 syllables, Z with 3 syllables
and W with 4 syllables.
These are shown in Table~\ref{tab:pw}
with fraction of the space actually used in Toki Pona.

@@

>>>>

\subsection{Statistics of the vocabulary}\label{sec:stat}
In~\cite{pije} are statistics about Toki Pona corpus.
This section focuses on the statistics of the vocabulary
and syntactic rules:
the letters, phonemes, word sizes,
possible combinations for words and sentences.
The statistics are in Appendix~\cite{listings},
and the next paragraph is an overview.
Python scripts were used to obtain the measurements
and are available at~\cite{tokipona}.

@@

Vowel and consonant frequencies are as shown in Table~\ref{freqLet}
for starting, end and internal position.
Comparison of vowels against consonant is in Table~\ref{tab:vcon}.
The high incidences e.g. of 'l' and 'a' favor
the naturalness of the language prosody
as it resembles child speech.

@@

26, 85 and 13 words have one, two and three syllables
(20.96, 68.54 and 10.48\% of all 124 tokens).
No official word have more than three syllables
(although proper names may be of any size).
In all the 124 tokens, there are 235 syllables (68 different).
Most often syllables are in
Table~\ref{syl}.
A complete list of the syllables and their frequencies (in different
positions) is in the \tttt{hsyls} variable of the \tttt{makeStatistics.py}
script.
A list of all words, grouped by their size in syllables,
and ordered alphabetically and by the number of letters,
is in \tttt{hlsyl\_\_}.

\input{syls}

Given the rules given in Section~\ref{phonology}
(of 14 letters, 5 vowels, (C)V(N) phonemes,
forbidden ji, wu, wo, ti, nn, nm),
46 words are possible with 1 syllable,
1886 with 2 syllables, 77326 with 3 syllables.
No middle syllable end with the nasal consonant 'n'.
These are shown in Table~\ref{tab:pw}
with fraction of the space actually used in Toki Pona.

@@

A toolbox for obtaining such statistics and sets
of words is in~\cite{tokipona}.
